MSstatsPTM: Statistical relative quantification of post-translational modifications in bottom-up mass spectrometry-based proteomics

### 2022-06-16 main/MSstatsPTM_Main.tex
Updated MSstats and MSstatsPTM stats sections

Consider a label-free LC-MS/MS experiment in the special case of a balanced design with $I$ conditions and $J$ biological replicates per condition. For simplicity, we assume that the experiment has no technical replicates, such that each biological replicate corresponds to an LC-MS/MS run.  For one protein, the PTM site is represented by $K$ spectral features (peptide ions, distinguished by their cleavage residues and charge states). The log-intensity (base 2) of Feature $k$, in Run $j$ of Condition $i$ is denoted by $y_{ijk}^{\ast}$. Conversely, the unmodified protein is represented by $L$ spectral features, and the log-intensity of Feature $l$ from the unmodified peptides in the same run is denoted by $y_{ijl}$. The features can be quantified as part of a same mass spectrometry run, or in separate enrichment and global proteome profiling runs.
Figure \ref{fig:data-structure}(a) schematically illustrates this data structure for one PTM site and the corresponding unmodified features of the protein, in a special case of 2 conditions and 2 biological replicates per condition \todo{OV: replace 'run' with 'bio rep' in the figure}. The number of modified and unmodified features varies across proteins. Some log-intensities may be outliers, and some spectral features can be missing. The population parameter of interest is the difference between the abundances of a PTM site in condition $1$ and condition $2$, denoted by $\mu_1^{\ast}$ and $\mu_2^{\ast}$ respectively. We are interested in testing the null hypothesis
\begin{equation}
\begin{aligned}
H_{0}: \Delta = \mu_{1}^{\ast} - \mu_{2}^{\ast} = 0 \\
H_{a}: \Delta = \mu_{1}^{\ast} - \mu_{2}^{\ast} \neq 0
\end{aligned}
\label{eq:conv_null_hyp}
\end{equation}
To test the hypothesis, a statistical model must represent the values in the table in Figure \ref{fig:data-structure}(a), and provide a model-based estimate of $\mu_{1}^{\ast} - \mu_{2}^{\ast}$, and its standard error.

@@

Many investigations perform differential abundance analysis of PTMs using two-sample $t$-test or its extensions. $t$-test can be performed without accounting for confounding with global protein abundance by taking as input intensities of individual features from modified peptides, denoted by $y_{ijk}^{\ast}$. Additionally, it can be extended to remove confounding by taking as input intensity ratios of modified and unmodified peptide features and comparing the mean abundance of a PTM site from one condition to another~\cite{Schwammle2015}\cite{THOMAS2020}\cite{Mertins:2013}. This is done by taking the sum of peak intensities in each run for modified and unmodified features separately and dividing the estimate of PTM abundance by the protein abundance estimate. This is equivalent to taking the difference of their log abundances. For condition $i$ and biological replicate $j$, the estimated quantity is denoted by $\hat{u}^{\ast}_{ij}$ and is given by Equation \ref{eq:adj_estimation}.

@@

These estimated ratios are then used as input to the two sample $t$-test, the output of which is the estimate of our quantity of interest $\hat{\Delta}$ as well as the standard error, $\widehat{SE(\hat{\Delta})}$. The statistical significance is determined by comparing the test statistic against the $t$ distribution, with degrees of freedom $df=2J-2$ in balanced designs. $t$-test assumes that the data follows a normal distribution, sampled from a representative portion of the total population, and that the conditions of interest have equal variance.

While simple, these approaches do not fully account for all sources of variations, and are not directly applicable to experiments with complex designs. These include experiments with comparisons of multiple conditions, acquisition in multiple batches, and unbalanced designs. Additionally, because the adjustment is done on the replicate, or run, level, there must be a matching number of replicates in both the modified and unmodified experimental designs. If the designs are different, this method is no longer directly applicable. Finally, while these approaches can be applied to experiments studying PTMs, there is not a self contained, straightforward implementation of the methods, making application challenging.

\medskip \noindent {\bf Limma} Modifications of $t$-test such as moderated $t$-test with Limma were also proposed~\cite{Ritchie_15a}\cite{Zhu}\cite{Chappell:2021}.

\todo{OV: could you also re-organize: (1) input, (2) model for the input, (3) model-based inference, (4) implementation, (5) good things and limitations. Limitation: limited in experimental design (at most two variance components, with strong assumptions)}

The inputs to Limma are the same as $t$-test. The intensities of individual features from modified peptides, $y_{ijk}^{\ast}$, can be used directly (in the presence of confounding with the unmodified protein). To remove confounding we transform the input by taking the intensity ratios of modified and unmodified peptide features, $\hat{u}^{\ast}_{ij}$, calculated in the same way as Equation \ref{eq:adj_estimation}. Limma then fits a linear model using the transformed inputs to estimate $\hat{\Delta}$. To estimate $\widehat{SE(\hat{\Delta})}$, Limma leverages empirical bayes moderation~\cite{Robbins1992} to moderate the model variance of the individual PTM using the shared global variance of all PTMs modeled in the experiment. For a single PTM $m$ with an estimated fold change $\hat{\Delta}_m$, the estimated variance is calculated as $s^{adj}_{m} = P(s_{m}| s_\Theta)$ where $s_{m}$ is the variance resulting from the linear model of PTM $m$ and $s_\Theta$ is the global variance across all PTMs. The statistical significance is determined using $\hat{\Delta}_m$ and $\widehat{SE(\hat{\Delta})}$ calculated using the adjusted variance $s^{adj}_{m}$.

@@

While a general improvement over $t$-test, there are still a number of limitations when using Limma. First, the method is limited in the types of experimental designs it can be applied to. The method is only directly applicable to experiments with at most two variance components, meaning that in more complex experiments it cannot account for all sources of variation. Additionally, it has the same issue as $t$-test with requiring the same experimental design for the modified and unmodified features, so that there are matching replicates in the ratio calculation. Finally, there is again no self contained implementation of the methods to PTMs, requiring manual transformation and application by the user at every step.

@@

\todo{OV: could you also re-organize: (1) input, (2) model for the input, (3) model-based inference, (4) implementation, (5) good things and limitations. Here good things are: complex designs, including multiple variance components; EB (for TMT); good summarization; implementation. Limitation: MSstats is not designed for protein-level adjustment of changes in PTM}

$MSstats$~\cite{Choi:2014} is designed for protein significance analysis and statistical relative quantification of proteins and peptides in global, targeted and data-independent proteomics. In terms of Figure \ref{fig:data-structure}, $MSstats$ was designed to estimate the difference in the unmodified protein in condition $1$ and condition $2$, denoted by $\mu_1$ and $\mu_2$ respectively. In other words it allows us to answer the new Null Hypothesis seen in Equation \ref{eq:msstats_null_hyp}.

@@

The package can be adjusted to analyze information on the peptide level, allowing us to estimate the convoluted Null Hypothesis in Equation \ref{eq:conv_null_hyp}. However, $MSstats$ currently includes no built in method to estimate the deconvoluted Null Hypothesis in Equation \ref{eq:null_hyp}.

$MSstats$ takes as input a list of log-transformed intensities of spectral features which are used to characterize the identified protein. For each protein, the feature log-intensities are expressed using a linear mixed model in consideration of the effects of condition, run, feature and interaction between run and feature. The model parameters are estimated using a split-plot approach, where the feature log-intensities are first summarized into a single value per protein per run in the subplot model, and the protein-level summaries are then used for the inference of the protein abundance. This approach allows the method to be extended to cases where the experimental design is unbalanced and additional sources of variation are present, overcoming the limitations in the previous sections. In the protein-level summarization, Tukey's median polish (TMP), a simple and robust procedure, is applied to iteratively fit a two-way additive model with the effects of run and feature, which in turn summarizes the log-intensities for each protein~\cite{Tukey:1977}. After summarization, the inference of the protein abundance in each condition is carried through fitting a model based on the family of linear mixed-effects models \cite{Bolker2009} \cite{Faraway:2006}. We leverage this statistical modeling and quantification workflow and apply it to PTMs, in addition to unmodified proteins.

@@

We propose a statistical method for detecting changes in PTMs that allows us to remove confounding with changes in the global protein, while being directly applicable to experiments with complex designs. To do this we leverage the existing methods in $MSstats$ and extend them to remove confounding between changes in the PTM and global protein. The method takes as input the spectral features of the PTM, $y_{ijk}^{\ast}$, and it's corresponding global protein, $y_{ijk}$. The features are summarized separately using the methods described previously for $MSstats$, resulting in run level values $\hat{y}_{1++}^{\ast}$ for the modified peptide and $\hat{y}_{1++}$ for the unmodified peptide. These summarized values are then used as the input to our models. Two separate linear mixed-effects models are fit, one using the summarized PTM values and another using the summarized global protein values. The models give estimates of the difference in log-abundance of the underlying PTM, $\hat{\Delta}_{PTM}$, and global protein, $\hat{\Delta}_{protein}$. Finally, the adjusted difference, $\hat{\Delta}$, is given by the difference in log-abundance of the PTM site, subtracted by the difference in log-abundance of the underlying protein.

@@

@@

As discussed above, the proposed approach can be extended to complex designs, including experiments with multiple sources of variation and unbalanced designs. Additionally, the approach can model experiments with separate designs for the PTM and global profiling run, i.e. differing number of runs and biological replicates, as long as the conditions of interest are the same. This extension is achieved through using the methods in $MSstats$, which can account for multiple sources of variation, and fully modeling both the PTM and global protein before adjustment. In terms of the workflow, first run level summarization is performed separately for both the PTM and unmodified protein using Tukey's Median Polish. Then a linear mixed effects model is fit to both the PTM and unmodified protein summarized features. The model takes into account all potential sources of variation, and restricted maximum likelihood (RML) is used to estimate the parameters of each model. Finally, once the parameters are estimated, we can combine the models using modified versions of Equations \ref{eq:prop_fc}, \ref{eq:prop_se}, and \ref{eq:prop_df}. Modeling the PTM and global protein separately and then combining them removes the need to match runs for adjustment, allowing us to model experiments with differing designs.

The log-fold change in the adjusted PTM abundance, $\hat{\Delta
}$, is now estimated by RML, and Equation \ref{eq:prop_fc} becomes 
\begin{equation}
\hat{\Delta} = (\hat{\mu}_{RML_i}^{\ast} - \hat{\mu}_{RML_{i'}}^{\ast}) - (\hat{\mu}_{RML_i} - \hat{\mu}_{RML_{i'}})
\end{equation}

The standard error of the estimate $\mathrm{SE}(\hat{\Delta})$ and degrees of freedom in Equations 3 and 4 are unchanged from the balanced case, although $\hat{\sigma}_{\gamma^\ast}$ and $\hat{\sigma}_{\gamma}$ are now estimated using RML.

@@

\begin{figure}[ht]
\centering
\includegraphics[scale=.6]{images/fig3.png}
\caption{
\todo{In 2a, it may be more meaningful to replace 'run' with 'bio rep'} Goals of PTM characterization, input to statistical analyses, and notation. (a) Schematic data representation, in a simplified case of two conditions and two biological replicates. Each PTM site is modeled and characterized separately, where a PTM is quantified with observed spectral features (boxes), distinguished by different charge states of a peptide. The feature \todo{log-}intensities of the PTM in Condition 1 are denoted by $y_1^\ast$ \todo{In the figure $y^*$ has 3 subscripts. Could you edit here as well?}. \todo{Log-intensities of } features corresponding to unmodified peptides in Condition 1 are denoted by $y_1$ \todo{In the figure $y$ has 3 subscripts. Could you edit here as well?}. Peptides can be fully cleaved (solid lines) and/or partially cleaved (dashed lines). \todo{OV: I am wondering whether we need Fig 2(b). The null hypothesis is specified in the text. I am not sure it is referenced anywhere. Maybe remove it for simplicity?} (b) PTM relative quantification of the population parameters by statistical inference, which makes use of the feature intensities to infer the underlying PTM abundance and protein abundance with an estimate of associated uncertainty. (c) Model-based testing for differential PTM abundance, which corrects for the underlying protein abundance with a cost of increased uncertainty about the estimate of difference between conditions.}
\label{fig:data-structure}
\end{figure}

>>>>

Consider a label-free LC-MS/MS experiment in the special case of a balanced design with $I$ conditions and $J$ biological replicates per condition. For simplicity, we assume that the experiment has no technical replicates, such that each biological replicate corresponds to an LC-MS/MS run.  For one protein, the PTM site is represented by $K$ spectral features (peptide ions, distinguished by their cleavage residues and charge states). The log-intensity (base 2) of Feature $k$, in Run $j$ of Condition $i$ is denoted by $y_{ijk}^{\ast}$. Conversely, the unmodified protein is represented by $L$ spectral features, and the log-intensity of Feature $l$ from the unmodified peptides in the same run is denoted by $y_{ijl}$. The features can be quantified as part of a same mass spectrometry run, or in separate enrichment and global proteome profiling runs.
Figure \ref{fig:data-structure}(a) schematically illustrates this data structure for one PTM site and the corresponding unmodified features of the protein, in a special case of 2 conditions and 2 biological replicates per condition. The number of modified and unmodified features varies across proteins. Some log-intensities may be outliers, and some spectral features can be missing. The population parameter of interest is the difference between the abundances of a PTM site in condition $1$ and condition $2$, denoted by $\mu_1^{\ast}$ and $\mu_2^{\ast}$ respectively. We are interested in testing the null hypothesis
\begin{equation}
\begin{aligned}
H_{0}: \Delta = \mu_{1}^{\ast} - \mu_{2}^{\ast} = 0 \\
H_{a}: \Delta = \mu_{1}^{\ast} - \mu_{2}^{\ast} \neq 0
\end{aligned}
\label{eq:conv_null_hyp}
\end{equation}
To test the hypothesis, a statistical model must represent the values in the table in Figure \ref{fig:data-structure}(a), and provide a model-based estimate of $\mu_{1}^{\ast} - \mu_{2}^{\ast}$, and its standard error.

@@

Many investigations perform differential abundance analysis of PTMs using two-sample $t$-test or its extensions~\cite{Schwammle2015}. These methods can be performed without accounting for convolution with global protein abundance by taking as input intensities of individual features from modified peptides, denoted by $y_{ijk}^{\ast}$. Additionally, they can be extended to remove convolution by taking as input intensity ratios of modified and unmodified peptide features and comparing the mean abundance of a PTM site from one condition to another~\cite{THOMAS2020}\cite{Mertins:2013}. This is done by taking the sum of peak intensities in each run for modified and unmodified features separately and dividing the estimate of PTM abundance by the protein abundance estimate. This is equivalent to taking the difference of their log abundances. For condition $i$ and biological replicate $j$, the estimated quantity is denoted by $\hat{u}^{\ast}_{ij}$ and is given by Equation \ref{eq:adj_estimation}.

@@

These estimated ratios are then used as input to the two sample $t$-test, the output of which is the estimate of our quantity of interest $\hat{\Delta}$ and the standard error, $\widehat{SE(\hat{\Delta})}$. The statistical significance is determined by comparing the test statistic against the $t$ distribution, with degrees of freedom $df=2J-2$ in balanced designs. $t$-test assumes that the data follows a normal distribution, sampled from a representative portion of the total population, and that the conditions of interest have equal variance.

While simple, these approaches do not fully account for all sources of variations, and are not directly applicable to experiments with complex designs. These include experiments with comparisons of multiple conditions, acquisition in multiple batches, and unbalanced designs. Additionally, because the adjustment is done on the replicate level, there must be a matching number of replicates in both the modified and unmodified experimental designs. If the designs are different, this method is no longer directly applicable. Finally, while these approaches can be applied to experiments studying PTMs, there is not a self contained, straightforward implementation of the methods, making application challenging.

\medskip \noindent {\bf Limma} Modifications of $t$-test such as moderated $t$-test with Limma were also proposed~\cite{Ritchie_15a}\cite{Zhu}\cite{Chappell:2021}. The inputs to the Limma based methods are the same as the inputs to $t$-test. The intensities of individual features from modified peptides, $y_{ijk}^{\ast}$, can be used directly (in the presence of confounding with the unmodified protein). To remove confounding we transform the input by taking the intensity ratios of modified and unmodified peptide features, $\hat{u}^{\ast}_{ij}$, calculated in the same way as Equation \ref{eq:adj_estimation}. Limma then fits a linear model using the transformed inputs to estimate $\hat{\Delta}$. To estimate $\widehat{SE(\hat{\Delta})}$, Limma leverages empirical bayes moderation~\cite{Robbins1992} to moderate the model variance of the individual PTM using the shared global variance of all PTMs modeled in the experiment. For a single PTM $m$ with an estimated fold change $\hat{\Delta}_m$, the estimated variance is calculated as $s^{adj}_{m} = P(s_{m}| s_\Theta)$ where $s_{m}$ is the variance resulting from the linear model of PTM $m$ and $s_\Theta$ is the global variance across all PTMs. The statistical significance is determined using $\hat{\Delta}_m$ and $\widehat{SE(\hat{\Delta})}$ calculated using the adjusted variance $s^{adj}_{m}$.

@@

While Limma offers improvements over $t$-test, the method still has a number of limitations. First, the method is limited in the types of experimental designs it can be applied to. The method is only directly applicable to experiments with at most two variance components, meaning that in more complex experiments it cannot account for all sources of variation. Additionally, it still requires the same experimental design for the modified and unmodified features so that the ratio calculation can be performed. Finally, there is no self contained implementation of the methods to PTMs, requiring manual transformation and application by the user.

@@

$MSstats$~\cite{Choi:2014} is designed for protein significance analysis and statistical relative quantification of proteins and peptides in global, targeted and data-independent proteomics. The package takes as input a list of intensities of individual features, denoted by $y_{ijk}$, used to characterize the identified protein. For each protein, the feature log-intensities are first summarized into a single value per protein per run in the subplot model. This is done using Tukey's median polish (TMP)~\cite{Tukey:1977}, a simple and robust procedure, which iteratively fits a two-way additive model with the effects of run and feature, summarizing them into a single value per run. These summarized values are then used as input to fit a linear model based on the family of linear mixed-effects models \cite{Bolker2009} \cite{Faraway:2006}. These methods are implemented in the packages $MSstats$ and $MSstatsTMT$~\cite{Huang:2020} and include functionalities for end to end quantification of proteins and differential analysis between experimental conditions.

The $MSstats$ analysis pipeline has a number advantages over simpler methods. First, the approach extends to many types of complex experimental design, including unbalanced designs, and additional sources of variation, such as fractionation. The methods include optional implementation of empirical bayes moderation, described previously, when applied to TMT experiments~\cite{Huang:2020}. Additionally, the summarization step includes options for missing value imputation, using an accelerated failure time (AFT) model~\cite{Wei:1992}, and has been shown to be more robust compared to using individual features~\cite{Choi:2014}.

While $MSstats$ is a powerful tool for the quantification of the global protein, it does not have built in tools to analyze PTMs. It is designed to estimate the difference in a protein between conditions, answering the Null Hypothesis

@@

It is not designed to remove confounding with the global protein meaning that it cannot directly address the Null Hypothesis in Equation \ref{eq:null_hyp}.

@@

We propose a statistical method for detecting changes in PTMs that allows us to remove confounding with changes in the global protein and is applicable to experiments with complex designs.   To do this we leverage the existing methods in $MSstats$ and extend them to remove confounding between changes in the PTM and global protein. The method takes as input the spectral features of the PTM, $y_{ijk}^{\ast}$, and it's corresponding global protein, $y_{ijk}$. The features are summarized separately using the methods described previously for $MSstats$, resulting in run level values $\hat{y}_{1++}^{\ast}$ for the modified peptide and $\hat{y}_{1++}$ for the unmodified peptide. These summarized values are then used as the input to our models. Two separate linear mixed-effects models are fit, one using the summarized PTM values and another using the summarized global protein values. The models give estimates of the difference in log-abundance of the underlying PTM, $\hat{\Delta}_{PTM}$, and global protein, $\hat{\Delta}_{protein}$. Finally, the adjusted difference, $\hat{\Delta}$, is given by the difference in log-abundance of the PTM site, subtracted by the difference in log-abundance of the underlying protein.

@@

\todo{DK: I'm not sure where this part should be. could move it to the first paragraph of this section, or in the extensions to complex design}
Finally, this approach allows us to model experiments with separate designs for the PTM and global profiling runs. This includes differing number of runs, biological replicates, or acquisition methods. This is achieved through fully modeling the PTM and global protein separately before adjusting. Because we are not combining on the replicate level, the number of replicates for each group do not need to match for adjustment. This makes the proposed method more flexible and applicable to a wider variety of experimental designs.

@@

As discussed above, the proposed approach can be extended to complex designs, including experiments with multiple sources of variation and unbalanced designs. This extension is achieved through utilizing the methods in $MSstats$ which, in complex cases, use restricted maximum likelihood (REML)~\cite{Kenward} to estimate the parameters of the models. Once the parameters are estimated, we can adjust the PTM model for changes in the global protein using modified versions of Equations \ref{eq:prop_fc}, \ref{eq:prop_se}, and \ref{eq:prop_df}.

The log-fold change in the adjusted PTM abundance, $\hat{\Delta
}$, is now estimated by REML, and Equation \ref{eq:prop_fc} becomes 
\begin{equation}
\hat{\Delta} = (\hat{\mu}_{RML_i}^{\ast} - \hat{\mu}_{RML_{i'}}^{\ast}) - (\hat{\mu}_{RML_i} - \hat{\mu}_{RML_{i'}})
\end{equation}

The standard error of the estimate $\mathrm{SE}(\hat{\Delta})$ and degrees of freedom in Equations 3 and 4 are unchanged from the balanced case, although $\hat{\sigma}_{\gamma^\ast}$ and $\hat{\sigma}_{\gamma}$ are now estimated using REML.

@@

\begin{figure}[ht]
\centering
\includegraphics[scale=.6]{images/fig3.png}
\caption{Goals of PTM characterization, input to statistical analyses, and notation. (a) Schematic data representation, in a simplified case of two conditions and two biological replicates. Each PTM site is modeled and characterized separately, where a PTM is quantified with observed spectral features (boxes), distinguished by different charge states of a peptide. The log-intensity of the PTM in Condition i, Run j, and feature k are denoted by $y_{ijk}^{\ast}$. Log-intensity of feature corresponding to unmodified peptide in Condition i, Run j, and feature k are denoted by $y_{ijk}$. Peptides can be fully cleaved (solid lines) and/or partially cleaved (dashed lines). (c) Model-based testing for differential PTM abundance, which corrects for the underlying protein abundance with a cost of increased uncertainty about the estimate of difference between conditions.}
\label{fig:data-structure}
\end{figure}
